Fluent statistical computing interfaces for biological data analysis

### 2020-02-23 Rmd/00-c-declaration.Rmd
fixing up declarations section

`r if (knitr::is_html_output()) '<!--'`

@@

\vspace*{2cm}\par\authorname

>>>>

---
output:
  pdf_document: default
  html_document: default
---
`r if (knitr::is_html_output()) '<!--'`

@@

This thesis includes 3 publications, two of which have been published and one 
which has not yet been submitted [@Lee2019; @Lee2020-ie] . As the core theme of my thesis is the development of software
interfaces for biological data analysis, and given the collaborative nature
of statistical computing and bioinformatics research; all of the included
papers in this thesis reflect and acknowledge the contributions of my 
co-authors, including my  primary supervisors Professor Dianne Cook (DC) in the 
Department of Econometrics and Business Statistics at Monash University
and Associate Professor Matthew Ritchie in the Epigenetics and Development 
Division at the Walter and Eliza Hall Institute. The following table details
the publications and my and my fellow co-authors contributions:

\begin{table}
\centering\footnotesize\tabcolsep=0.12cm
\begin{tabular}{|p{1cm}|p{2cm}|p{1.5cm}|p{3.5cm}|p{3.5cm}|p{1.5cm}|}
\hline
\RaggedRight\textbf{Thesis Chapter}  & \RaggedRight\textbf{Publication Title}  & \RaggedRight\textbf{Status (published, in press, accepted or returned for revision)}  & \RaggedRight\textbf{Nature and} {\%} \RaggedRight\textbf{of student contribution} & \RaggedRight\textbf{Co-author name(s) Nature and} {\%} \RaggedRight\textbf{of Co-author’s contribution} & \RaggedRight\textbf{Co-author(s), Monash student Y/N} \\ \hline
2 & plyranges: a grammar of genomic data transformation & Published  & 70\%. Concept, software development, data analysis, and manuscript writing & (1) Dianne Cook, Concept and manuscript revision 10\% (2) Michael Lawrence, Concept and software development  20\% & \multicolumn{1}{c|}{N}  \\
\hline
3 & Fluent genomics with \textit{plyranges} and \textit{tximeta} & Published (awaiting peer review) & 60\%. Concept, software development, data analysis and manuscript writing  & (1) Michael Lawrence, manuscript feedback and editing 5\% (2) Michael I Love, Concept, data analysis and manuscript writing 35\% & \multicolumn{1}{c|}{N} \\
\hline
5 & Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings & To be submitted & 80\%. Concept, software development, and manuscript writing & (1) Dianne Cook, Concept and manuscript writing 15\% & \multicolumn{1}{c|}{N} \\
\hline
\end{tabular}
\end{table}

\FloatBarrier

Note that chapter \@ref(ch:intron), reflects my  contribution (in the form of a 
software package and data analysis) to the submitted manuscript entitled 
_Covering all your bases: incorporating intron signal from RNA-seq data_
[@Lee2019-mf] and is joint work under the supervision of Dr Charity Law
and Associate Professor Matthew Ritchie.

\textbf{Student name:} Stuart Andrew Lee

\textbf{Student signature:}

```{r signature, out.width = "20%", fig.align = "left", eval = file.exists("img/signature.png")}
knitr::include_graphics("img/signature.png")
```

### 2020-02-23 Rmd/05-embeddings.Rmd
fixing up declarations section

High dimensional data is increasingly prevalent in the natural sciences and
beyond but presents a challenge to the analyst in terms of both data cleaning /
preprocessing and visualisation. Methods to embed data from a high-dimesional
space into a low-dimensional one now form a core step of the data analysis
workflow where they are used to ascertain hidden structure and denoise data for
downstream analysis (thereby nullifying the 'curse of dimensionality').

@@

Non-linear methods generally perform preprocessing on the high-dimensional data
such as generating a neighborhood graph and perform transformations on the
preprocessed form. We restrict our attention to three methods that are commonly
used in high-throughput biology: t-distributed stochastic neighbor embedding
(t-SNE), uniform manifold alignment and projection (UMAP), and potential of
heat-diffusion for affinity-based transition embedding (PHATE). The t-SNE
algorithm estimates the similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. The resulting configuration is
the one that minimizes the Kulback-Leibler divergence between the two
distributions. A recent theoretical contribution by @Linderman2019-dq proved
that t-SNE can recover spherical and well separated cluster shapes, and
proposed new approaches for tuning the optimisation parameters. It is a known
problem that t-SNE can have trouble recovering global structure and that
configurations can be highly dependent on how the algorithm is initialised and
parameterised [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
neighbor graph. Its main differences are that it uses a different cost function
(cross entropy) which is optimized using stochastic gradient descent and
defines a different kernel for similiarites in the low dimensional space. Due
to it's computational speed it's possible to generate UMAP embeddings in more
than three dimensions. Finally, PHATE, inspired by diffusion maps, is based on
estimating an affinity matrix via a distance matrix and k-nearest neighbors
graph. The algorithm denoises estimated distances in high dimensional space via
transforming the affinity matrix into a Markov transition probaility matrix and
diffusing this matrix over a fixed number of time steps. Then the diffused
probabilities are transformed once more to construct a distance matrix, and
multidimensional scaling is performed to construct a 2d embedding for
visualization.

As part of a visualization workflow, it's important to consider the perception
and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
2D scatter plots were mostly suffucient for detecting class seperation, however
they noted that often multiple embeddings were required. For the task of
cluster identification, @Lewis2012-ai showed experimentally that novice users
of non-linear embedding techniques were more likely to consider clusters of
points on a 2d scatter plot to be the result of a spurious embedding compared
to advanced users who were aware of the inner workings of the embedding
algorithm.

@@

While there has been much work on the algorithmic details of the aformentioned
embedding methods, there has been relatively few tools designed to assist users
to interact with these techniques and assist them in making comparsions between
embeddings and performing the aforementioned cluster orientation tasks. Several
interactive interfaces have been proposed for evaluating or using embedding
techniques: the Sleepwalk interface provides a click and highlight
visualisation for colouring points in an embedding according to their distance
in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
algorithm, that allows users to modified optimisation parameters in real-time.
Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
PCA live in the browser and provides interactions to color points, and
highlights nearest neighbors [@Smilkov2016-hp].

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and somewhat poorly defined problem. For non-linear
methods, there are a lot of parameters to explore that can have an effeft on
the resulting visualisation and interpretation. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. We propose a more pragmatic workflow inspired by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with
cluster orientation tasks. This workflow is incorporated into an R package
called `liminal` (Available: https://github.com/sa-lee/liminal).

@@

To investigate latent structure and the shape of a high dimensional dataset, a
tour can be run without the use of an external embedding. It is often useful to
first run principal components on the input as an intial dimension reduction
step, and then tour a subset of those components instead, i.e. by selecting
them from a scree plot. The default tour layout is a scatter plot with an axis
layout displaying the magnitude and direction of each basis vector. Since the
tour is dynamic, it is often useful to be able to pause and highlight a
particular view. In our interface, brushing will pause the tour path, allowing
users to identify 'interesting' projections. The domain of the axis scales from
running a tour is called the half range, and is computed by rescaling the input
data onto hyperdimensional unit cube. We bind the half range to a mouse wheel
event, allowing a user to pan and zoom on the tour view dynamically. This is
useful for peeling back dense clumps of points to reveal structure.

@@

We combine the tour view in a side by side layout with a scatter
plot view from an embedding algorithm. The views are linked; analysts can brush a group of points in the embedding view, and see its shape in the tour view.
This can be used for either manually identifying clusers, or verifying the location of clusters relative to other groupings. Consequently, simple linked
brushing can alleviate a downfall of non-linear methods such as UMAP or t-SNE
that break hierachial structure of clusterings. Aside from simple
linked brushing, we can also pose more complex queries via interrogating
the neighbourhood graph. Using a linked neighbourhood brush, we can 
visually investigate the nearest neighbour relationships in the 
high-dimensional space via brushing in the embedding view. The user can select
the number of nearest neighbours directly, and modify the distance metric 
used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
can be used to pose queries in either the tour view or the embedding view;
interface controls allow these brushes to combine using logical operators
such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
single point highlights or colorings, allowing analysts to get a more holistic
view of the effect of an embedding algorithm.

### Making comparsions

@@

### Case Study 3: largeish single cell data

>>>>

High dimensional data is increasingly prevalent in the natural sciences and
beyond but presents a challenge to the analyst in terms of both data cleaning /
pre-processing and visualisation. Methods to embed data from a high-dimensional
space into a low-dimensional one now form a core step of the data analysis
workflow where they are used to ascertain hidden structure and de-noise data for
downstream analysis (thereby nullifying the 'curse of dimensionality').

@@

Non-linear methods generally perform pre-processing on the high-dimensional data
such as generating a neighborhood graph and perform transformations on the
pre-processed form. We restrict our attention to three methods that are commonly
used in high-throughput biology: t-distributed stochastic neighbor embedding
(t-SNE), uniform manifold alignment and projection (UMAP), and potential of
heat-diffusion for affinity-based transition embedding (PHATE). The t-SNE
algorithm estimates the similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. The resulting configuration is
the one that minimizes the Kullback-Leibler divergence between the two
distributions. A recent theoretical contribution by @Linderman2019-dq proved
that t-SNE can recover spherical and well separated cluster shapes, and
proposed new approaches for tuning the optimisation parameters. It is a known
problem that t-SNE can have trouble recovering global structure and that
configurations can be highly dependent on how the algorithm is initialised and
parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
neighbor graph. Its main differences are that it uses a different cost function
(cross entropy) which is optimized using stochastic gradient descent and
defines a different kernel for similarities in the low dimensional space. Due
to it's computational speed it's possible to generate UMAP embeddings in more
than three dimensions. Finally, PHATE, inspired by diffusion maps, is based on
estimating an affinity matrix via a distance matrix and k-nearest neighbors
graph. The algorithm de-noises estimated distances in high dimensional space via
transforming the affinity matrix into a Markov transition probability matrix and
diffusing this matrix over a fixed number of time steps. Then the diffused
probabilities are transformed once more to construct a distance matrix, and
multidimensional scaling is performed to construct a 2d embedding for
visualization.

As part of a visualization workflow, it's important to consider the perception
and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
2D scatter plots were mostly sufficient for detecting class separation, however
they noted that often multiple embeddings were required. For the task of
cluster identification, @Lewis2012-ai showed experimentally that novice users
of non-linear embedding techniques were more likely to consider clusters of
points on a 2d scatter plot to be the result of a spurious embedding compared
to advanced users who were aware of the inner workings of the embedding
algorithm.

@@

While there has been much work on the algorithmic details of the aforementioned
embedding methods, there has been relatively few tools designed to assist users
to interact with these techniques and assist them in making comparisons between
embeddings and performing the aforementioned cluster orientation tasks. Several
interactive interfaces have been proposed for evaluating or using embedding
techniques: the Sleepwalk interface provides a click and highlight
visualisation for colouring points in an embedding according to their distance
in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
algorithm, that allows users to modified optimisation parameters in real-time.
Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
PCA live in the browser and provides interactions to color points, and
highlights nearest neighbors [@Smilkov2016-hp].

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and somewhat poorly defined problem. For non-linear
methods, there are a lot of parameters to explore that can have an effect on
the resulting visualisation and interpretation. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. We propose a more pragmatic workflow inspired by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with
cluster orientation tasks. This workflow is incorporated into an R package
called `liminal` (Available: https://github.com/sa-lee/liminal).

@@

To investigate latent structure and the shape of a high dimensional dataset, a
tour can be run without the use of an external embedding. It is often useful to
first run principal components on the input as an initial dimension reduction
step, and then tour a subset of those components instead, i.e. by selecting
them from a scree plot. The default tour layout is a scatter plot with an axis
layout displaying the magnitude and direction of each basis vector. Since the
tour is dynamic, it is often useful to be able to pause and highlight a
particular view. In our interface, brushing will pause the tour path, allowing
users to identify 'interesting' projections. The domain of the axis scales from
running a tour is called the half range, and is computed by rescaling the input
data onto hyper-dimensional unit cube. We bind the half range to a mouse wheel
event, allowing a user to pan and zoom on the tour view dynamically. This is
useful for peeling back dense clumps of points to reveal structure.

@@

We combine the tour view in a side by side layout with a scatter
plot view from an embedding algorithm. The views are linked; analysts can brush a group of points in the embedding view, and see its shape in the tour view.
This can be used for either manually identifying clusters, or verifying the location of clusters relative to other groupings. Consequently, simple linked
brushing can alleviate a downfall of non-linear methods such as UMAP or t-SNE
that break hierarchical structure of clusters. Aside from simple
linked brushing, we can also pose more complex queries via interrogating
the neighbourhood graph. Using a linked neighbourhood brush, we can 
visually investigate the nearest neighbour relationships in the 
high-dimensional space via brushing in the embedding view. The user can select
the number of nearest neighbours directly, and modify the distance metric 
used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
can be used to pose queries in either the tour view or the embedding view;
interface controls allow these brushes to combine using logical operators
such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
simple color highlights, allowing analysts to get a more holistic
view of the effect of an embedding algorithm.

### Making comparisons

@@

### Case Study 3: largish single cell data
